RationalGRL: A Framework for Argumentation and Goal Modeling

### 2018-05-02 Response to reviewers.txt

######## EDITOR ########

===== COMMENT #1
Please ensure also that your manuscript shows clearly how your work relates to relevant works as they may have been published in the Requirements Engineering or presented at the Requirements Engineering series of conferences.

===== REPLY #1
Currently 6 of the 44 references in the article come from either REJ or the RE conference [1,2,3,4,5,6]. We have searched through the journal and conference database for more references, and we found 4 more references that are closely related to our work [7,8,9,10]. We have added a discussion of these articles to the related work section, at the end of the paragraph "Requirements Engineering". This means we currently have 10 references from the Requirements Engineering Journal or presented at the Requirements Engineering series of conferences.

@@

######## REVIEWER #1 ########

We thank the reviewer for the insightful comments. Especially the comment about the missing validation was useful since it motivated us to perform an small-scale in-house user study, which allowed us to strengthen our claims for the usefulness and usability of our framework. We’ve replied to each separate comment below.

===== COMMENT #1
The integrated meta model looks reasonable but I have some doubts about the inclusion of "critical question" in the schema. Critical questions are something that are used to identify arguments and attack links, but  I don't see them as a concept to be included in the meta model. They are part of the methodology used to build the goal model rather concepts of the language.

===== REPLY #1
We thank the reviewer for pointing this out. We agree critical questions should not be part of the metamodel, and we have removed them accordingly. We also have updated the text accordingly.

@@

===== COMMENT #2
The proposed algorithms are very simple and they do not add much to the overall presentation, I think they can be summarized in the section describing the tool.

===== REPLY #2
We agree that many algorithms are similar, and we have therefore shortened the sections containing these algorithms: sections 5.4.1 to 5.4.3 now all contain one example of the specific type of algorithm instead of two or more.

@@

===== COMMENT #3
The methodology and the tools looks quite preliminary and without a real validation of the approach. I think this is the weakest part of the paper. Authors should demonstrate the applicability, scalability and usefulness of their approach in the real word. The tool and the modeling language should be used and evaluated by practitioners while they cope with real requirements engineering problems and complexity issues.

===== REPLY #3
We agree with the reviewer that a real validation of the approach is a necessary next step. However, we also would like to point out that an industrial strength case study is a significant amount of work, and would most likely have enough new insights for a separate paper. To give an example, it took Estrada et al. [11] about 9 months to conduct the case study.

Since the main focus of our article is to explain the noval RationalGRL framework in detail, we feel that adding such an extensive study to the current article may be too much content. Still, we agree that having no validation at all is limiting, so we have performed two small-scale user studies in order to provide an initial validation of our framework. These two studies are as follows:
1) We evaluated the RationalGRL tool with two Enterprise Architects from a large bank in the Netherlands. The architects found the RationalGRL models very useful in that they could visualize the arguments and were easier for them to understand the links between the rationale, and the resulting goal model elements.
2) We performed an empirical study amongst ==> TODO(marcvanzee): ADD NUMBER HERE  <== participants, each with a computer science background.

We have added a section "Validation" (Section 7) where we describe the results of the validation.

@@

######## REVIEWER #2 ########

We thank the reviewer for reading our article in detail, and providing many insightful and useful comments. We’ve replied to each separate comment below.

===== COMMENT #1
I think that the relatively simple argument structures are appealing in the sense that the overhead for practical application may be tolerable. However, this remains to be shown yet in an industrial environment.

===== REPLY #1
We have added a section "Validation" (Section 7) where we describe the results of a validation. Please see REPLY #3 to REVIEWER #1 for more details.

@@

===== COMMENT #2
The statement on actually using a reduced GRL should be made in the Abstract and Introduction already, not just in a footnote later.

===== REPLY #2
We have included in the abstract and introduction that we use a simplified version. Note that all the core elements of GRL are included but, for example, we have included only one level of contribution strength, where GRL contains different levels of positive and negative contribution. This simplification does not in any way influence the case study or any of the other results. Furthermore, extending the GRL part of RationalGRL to include the full set of GRL elements would not change the essence of our formalism.

===== COMMENT #3
I think that the paper should compare the proposed argumentation structure at least with the beliefs in GRL. In general, how does this argumentation approach really differ from one based on beliefs and, in particular, belief revision? Assume that an argument is a belief of some stakeholder revised by another argument, which is a belief of some other stakeholder?

===== REPLY #3
We have added a comparison between arguments in RationalGRL and beliefs in GRL at the end of the related work section, in the paragraph "Beliefs in GRL".

The comparison between argumentation and belief revision is interesting, especially from a theoretical point of view. In the area of knowledge representation, obtaining correspondences between beliefs revision and the dynamics of argumentation theory is a popular topic of research [12,13,14,15]. However, since studying relations between argumentation and belief revision does not seem directly related to our current framework and seems more interesting from a theoretical point of view, we have left out the theoretical comparison. Instead, we chose to compare it with beliefs in GRL.

@@

===== COMMENT #8
Regarding terminology, I wonder about "valid" GRL models here, which seem to mean that certain criteria are fulfilled. Validation would mean that a goal model makes sense in the real world as having the "right" goals and relationships between them.

===== REPLY #8
We agree that the term “valid” is confusing in this respect, since we do not want to express that the goal model makes sense in the real world. Instead, we’d like to express that certain criteria are fulfilled. Therefore, we’ve renamed “valid” to “consistent”. 
TODO(marcvanzee): Implement this renaming in the article.

===== COMMENT #9
"Verification" in 6.2 appears to stand for evaluation. This cannot be "verified".

===== REPLY #9
We have renamed “verification” to “demonstration” and slightly reworded the sentence.

@@

===== COMMENT #10
I also do not think that Figure 16 shows a "methodology". It does not even look like a "method".

===== REPLY #10
We agree that the terminology is wrong. We have renamed the methodology to "The RationalGRL development process", since it defines the steps on how to create a RationalGRL model. We have renamed this throughout the entire article.

@@

===== COMMENT #11
please reconsider the very first sentences in the Abstract and Introduction each, especially the latter statement on RE

===== REPLY #11
We agree with the reviewer that the two mentioned sentences are confusing. We have improved this as follows:
- We have rewritten the first sentence of the abstract as follows: "Goal-oriented requirements modeling approaches aim at capturing intentions of various stakeholders involved in the development of an information system and refining them into low level goals and tasks."
- We have deleted the first sentence of the introduction, since we feel that it is not necessary to explain the definition of Requirements Engineering in the Requirements Engineering Journal.

@@

===== COMMENT #12
Finally, you should run a spell checker, have a careful proof reading of grammar, and make sure that there is no unresolved reference to a figure in a submitted manuscript.

===== REPLY #11
We’ve ran a spell-checker and proof-read the paper carefully. There was one unresolved reference to a figure, which a corrected. Thank you for pointing this out.

######## REFERENCES ########
[1] Y. Elrakaiby, A. Ferrari, P. Spoletini, S. Gnesi, and B. Nuseibeh. Using argumentation to explain ambiguity in requirements elici- tation interviews. In Proceedings of the 25th IEEE International Requirements Engineering Conference (RE 2017), pages 51–60. IEEE Press, 2017.
[2] I. Jureta, S. Faulkner, and P. Schobbens. Clear justification of mod- eling decisions for goal-oriented requirements engineering. Re- quirements Engineering, 13(2):87–115, May 2008.
[3] Y. Liu, Y. Su, X. Yin, and G. Mussbacher. Combined goal and feature model reasoning with the user requirements notation and jucmnav. In Requirements Engineering Conference (RE), 2014 IEEE 22nd International, pages 321–322. IEEE, 2014.
[4] P. K. Murukannaiah, A. K. Kalia, P. R. Telangy, and M. P. Singh. Resolving goal conflicts via argumentation-based analysis of com- peting hypotheses. In Proceedings of the 23rd International Re- quirements Engineering Conference (RE 2015), pages 156–165. IEEE, 2015.
[5] M. van Zee, F. Bex, and S. Ghanavati. Rationalization of Goal Models in GRL using Formal Argumentation. In Proceedings of the 23rd International Requirements Engineering Conference (RE’15), pages 220–225. IEEE Press, August 2015
[6] E. S. K. Yu. Towards Modeling and Reasoning Support for Early-Phase Requirements Engineering. In Proceedings of the 3rd IEEE International Symposium on Requirements Engineering, pages 226–235, 1997.

[7] S. R. Smith. Constructing and revising formal arguments in re- quirements engineering. Requir. Eng., 2(4):199–216, Dec. 1997.
[8] J.Lockerbie,N.A.M.Maiden,J.Engmann,D.Randall,S.Jones, and D. Bush. Exploring the impact of software requirements on system-wide goals: a method using satisfaction arguments and i* goal modelling. Requirements Engineering, 17(3):227–254, 2012.
[9] J. Horkoff and E. Yu. Interactive goal model analysis for early requirements engineering. Requirements Engineering, 21(1):29– 61, 2016.
[10] J. Hassine and D. Amyot. A questionnaire-based survey method- ology for systematically validating goal-oriented models. Require- ments Engineering, 21(2):285–308, 2016.

[11] Estrada, H., Rebollar, A. M., Pastor, O., & Mylopoulos, J. (2006, June). An empirical evaluation of the i* framework in a model-based software generation environment. In International Conference on Advanced Information Systems Engineering (pp. 513-527). Springer, Berlin, Heidelberg.

[12] Falappa, Marcelo Alejandro, Gabriele Kern-Isberner, and Guillermo Ricardo Simari. "Belief revision and argumentation theory." Argumentation in artificial intelligence. Springer, Boston, MA, 2009. 341-360.
[13] Liao, Beishui, Li Jin, and Robert C. Koons. "Dynamics of argumentation systems: A division-based method." Artificial Intelligence 175.11 (2011): 1790-1814.
[14] Dix, Jürgen, et al. "Belief change and argumentation in multi-agent scenarios (Dagstuhl seminar 13231)." Dagstuhl Reports. Vol. 3. No. 6. Schloss Dagstuhl-Leibniz-Zentrum fuer Informatik, 2013.
[15] Paglieri, Fabio, and Cristiano Castelfranchi. "Revising beliefs through arguments: Bridging the gap between argumentation and belief revision in MAS." International Workshop on Argumentation in Multi-Agent Systems. Springer, Berlin, Heidelberg, 2004.

### 2018-05-02 abstract.tex
Addressed some more comments by reviewers

\begin{abstract}
Goal modeling languages capture the relations between an information system and its environment using high-level goals and their relationships with lower level goals and tasks. The process of constructing a goal model usually involves discussions between a requirements engineer and a group of stakeholders. While it is possible to capture part of this discussion process in a goal model, for instance by specifying alternative solutions for a goal, not all of the arguments can be found back in the resulting model. For instance, the discussion on whether to accept or reject a certain goal and the ultimate rationale for acceptance or rejection cannot be captured in current goal modeling languages. Based on a case study in which stakeholders discuss requirements for a Traffic Simulator, we apply argumentation techniques from artificial intelligence to a goal modeling approach. We combine a simplified version of a traditional goal modeling approach, the Goal-oriented Requirements Language (GRL), with a formal Practical Reasoning Argument Scheme (PRAS) for reasoning about goals into a new framework (RationalGRL). RationalGRL provides a methodology, formal semantics and tool support to capture the discussions and outcomes of the argumentation process that leads to a goal model.

>>>>

\begin{abstract}
Goal-oriented requirements modeling approaches aim at capturing intentions of various stakeholders involved in the development of an information system and refining them into low level goals and tasks. The process of constructing a goal model usually involves discussions between a requirements engineer and a group of stakeholders. While it is possible to capture part of this discussion process in a goal model, for instance by specifying alternative solutions for a goal, not all of the arguments can be found back in the resulting model. For instance, the discussion on whether to accept or reject a certain goal and the ultimate rationale for acceptance or rejection cannot be captured in current goal modeling languages. Based on a case study in which stakeholders discuss requirements for a Traffic Simulator, we apply argumentation techniques from artificial intelligence to a goal modeling approach. We combine a simplified version of a traditional goal modeling approach, the Goal-oriented Requirements Language (GRL), with a formal Practical Reasoning Argument Scheme (PRAS) for reasoning about goals into a new framework (RationalGRL). RationalGRL provides a formal semantics and tool support to capture the discussions and outcomes of the argumentation process that leads to a goal model. We also define the RationalGRL development process to create a RationalGRL model.

### 2018-05-02 discussion.tex
Addressed some more comments by reviewers

Software design and requirements engineering are very closely related \cite{nuseibeh2001weaving} and hence the insights from the DR literature are directly applicable to RE. The RationalGRL framework essentially incorporates the core ideas from DR into goal-oriented requirements engineering by explicitly including arguments for and against the various options into the goal model, and by proposing a methodology and critical questions that encourage reflection when thinking about the possible goals and functionality of a system.

@@

The GAM process is essentially a generic, high-level process for problem solving, not specifically tailored towards goal modeling. In this sense, the RationalGRL methodology in Figure~\ref{fig:rationalgrl-methodology} can be seen as a more specific version of GAM explicitly meant for goal modeling. The argument schemes and critical questions in RationalGRL provide clearer handles for goal modeling (cf. requirement 4 of our framework, Section~\ref{sect:introduction}). Furthermore, the RationalGRL framework adds full tool support for the goal modeling and reasoning process, which GAM lacks. GAM does contain more specific ways of dealing with different types of clarity problems; in RationalGRL clarification is captured as a single critical question (CQ12, Table~\ref{table:argument-schemes}). However, as argued above further critical questions can easily be added to RationalGRL if needed.

@@

@@

\paragraph{Empirical validation}
We have performed a case study and found more than 200 incidences of arguments and questions. However, what is still lacking is further empirical investigation of whether, and how, the tools and techniques provided by the RationalGRL framework really improve early phase requirements engineering. The complexity of the domain makes focused experiments difficult, but not impossible. For example, Murakannaiah \emph{et al.}~\cite{murukannaiah2015} provided a realistic but bounded problem for their subjects to solve, and similar experiments are imaginable to test RationalGRL: have two groups solve a goal modeling problem, one using only GRL and one using RationalGRL, and examine how they perform. Naturally, the type of problem will greatly influence the outcome. For a simple problem which two people have to solve in an hour, RationalGRL will most likely be of little benefit. However, a more complex problem on which multiple persons have to work asynchronously might benefit from the extra tools offered by RationalGRL. Furthermore, the outcome also depends on which part of the set of tools and techniques offered by RationalGRL is tested: the methodology, the tool, or a list of critical questions to serve as reminders during the goal modeling process (cf. \cite{SchriekEtal2016}).

@@

One of our main goals is to provide means for RE practitioners to capture the underlying arguments of goal models by using the RationalGRL framework (\textbf{requirement 4}). To this end, we propose a methodology, which consists of developing goal models and posing arguments based on schemes in an integrated way. Finally, we have implemented the RationalGRL tool, a web-based prototype\footnote{\url{http://www.rationalgrl.com}}, for modeling argument schemes and critical questions and for reasoning about goal models with respect to them (\textbf{requirement 5}).

>>>>

Software design and requirements engineering are very closely related \cite{nuseibeh2001weaving} and hence the insights from the DR literature are directly applicable to RE. The RationalGRL framework essentially incorporates the core ideas from DR into goal-oriented requirements engineering by explicitly including arguments for and against the various options into the goal model, and by proposing a development process and critical questions that encourage reflection when thinking about the possible goals and functionality of a system.

@@

The GAM process is essentially a generic, high-level process for problem solving, not specifically tailored towards goal modeling. In this sense, the RationalGRL Development Process in Figure~\ref{fig:rationalgrl-methodology} can be seen as a more specific version of GAM explicitly meant for goal modeling. The argument schemes and critical questions in RationalGRL provide clearer handles for goal modeling (cf. requirement 4 of our framework, Section~\ref{sect:introduction}). Furthermore, the RationalGRL framework adds full tool support for the goal modeling and reasoning process, which GAM lacks. GAM does contain more specific ways of dealing with different types of clarity problems; in RationalGRL clarification is captured as a single critical question (CQ12, Table~\ref{table:argument-schemes}). However, as argued above further critical questions can easily be added to RationalGRL if needed.

@@

\paragraph{Beliefs in GRL and Belief Revision} As mentioned already in the introduction, it is possible in GRL to rationalize models
through so-called \emph{belief} elements. These beliefs are elements that can be attached to an IE with a \emph{belief link}. The content of a belief is a textual description, which can explain, for instance, the rationale for including an IE in the resulting goal model. There are several important differences between GRL beliefs and RationalGRL arguments.

First of all, beliefs in GRL are not used in any of the algorithms to compute the satisfiability of the IEs. This means that a GRL belief can be understood as metadata that can be attached to an element, but it does not have semantics. Argument in RationalGRL, in contrast, play an important role in forming the resulting goal model. If an IE is defeated by an argument, it is not part of the resulting goal model.

Secondly, arguments in RationalGRL can not only provide rationale for IEs in the resulting goal model, but they can also provide rational for those that are rejected, and they can be attached to links as well. This means that Rational arguments are strictly more expressive than GRL beliefs.

Finally, the formation of arguments and counterarguments in RationalGRL is guided by critical questions, which stimulate discussions between stakeholders. This is not possible for GRL beliefs.

@@

\paragraph{Empirical validation}
We have performed a case study and found more than 200 incidences of arguments and questions. However, what is still lacking is further empirical investigation of whether, and how, the tools and techniques provided by the RationalGRL framework really improve early phase requirements engineering. The complexity of the domain makes focused experiments difficult, but not impossible. For example, Murakannaiah \emph{et al.}~\cite{murukannaiah2015} provided a realistic but bounded problem for their subjects to solve, and similar experiments are imaginable to test RationalGRL: have two groups solve a goal modeling problem, one using only GRL and one using RationalGRL, and examine how they perform. Naturally, the type of problem will greatly influence the outcome. For a simple problem which two people have to solve in an hour, RationalGRL will most likely be of little benefit. However, a more complex problem on which multiple persons have to work asynchronously might benefit from the extra tools offered by RationalGRL. Furthermore, the outcome also depends on which part of the set of tools and techniques offered by RationalGRL is tested: the development process, the tool, or a list of critical questions to serve as reminders during the goal modeling process (cf. \cite{SchriekEtal2016}).

@@

One of our main goals is to provide means for RE practitioners to capture the underlying arguments of goal models by using the RationalGRL framework (\textbf{requirement 4}). To this end, we propose a development process, which consists of developing goal models and posing arguments based on schemes in an integrated way. Finally, we have implemented the RationalGRL tool, a web-based prototype\footnote{\url{http://www.rationalgrl.com}}, for modeling argument schemes and critical questions and for reasoning about goal models with respect to them (\textbf{requirement 5}).

### 2018-05-02 introduction.tex
Addressed some more comments by reviewers

Requirements Engineering (RE) is an approach to assess the role of a future information system within its environment. An important goal in RE is to produce a consistent and comprehensive set of requirements covering different aspects of the system, such as general functional requirements, operational environment constraints, and so-called non-functional requirements such as security and performance.

Among the ``early-phase'' requirements engineering activities are those that consider how the intended system should meet organizational goals, why it is needed, what alternatives may exist, what the implications of the alternatives are for different stakeholders, and how the interests and concerns of stakeholders might be addressed~\cite{yu1997towards}. These activities fall under the umbrella of goal modeling. There are a large number of established RE methods using goal models in the early stage of requirements analysis (overviews can be found in~\cite{kavakliL05,van2001goal}). Several goal modeling languages have been developed in the last two decades as well. The most popular ones include $i*$~\cite{yu1997towards}, Keep All Objects Satisfied (KAOS)~\cite{van2008requirements}, the NFR framework~\cite{chung2012non}, \textsc{Tropos}~\cite{giorgini2005goal}, the Business Intelligence Model (BIM)~\cite{horkoff2014strategic}, and the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}.

@@

\begin{enumerate}
\item 
The argumentation techniques must capture the actual discussions of the stakeholders or designers in the early requirements engineering phase.
\item 
The framework must have formal traceability links between elements of the goal model and their underlying arguments.
\item 
Using these traceability links, it must be possible to compute the effect of changes in the underlying arguments about the goal model, and vice versa.
\item 
There must be a methodology to guide practitioners in using the framework.
\item 
The framework must have tool support.
\end{enumerate}

@@

In order to support practitioners in using the RationalGRL framework, we propose a methodology, which consists of developing goal models and posing arguments based on schemes in an integrated way. We implement a web-based tool for the RationalGRL framework as well, which works on any modern browser and is developed in Javascript.\footnote{The tool and source code can be found at \rationalgrlurl.} The tool is open-source, and designed such that new argument schemes and critical questions can be added easily.

The rest of this article is organized as follows. Section~\ref{sect:background} introduces our running example and briefly discusses the basics of GRL and argumentation about goals using argument schemes. Section~\ref{sect:gmas} contains the case study and an explanation of how we obtained an initial set of argument schemes and critical questions by coding transcripts from discussions about an information system. Section~\ref{sect:overview} provides an overview of the RationalGRL framework and the RationalGRL metamodel, and examples from the case study that illustrate the framework and its language. In Section~\ref{sect:formalframework}, we provide formal semantics for GRL and RationalGRL, show how RationalGRL models can be translated to GRL models and vice versa, and we develop various algorithms that change a RationalGRL model according to an argument scheme or a critical question. Section~\ref{sect:methodology+tool} discusses the RationalGRL methodology and explains various features of the tool we developed. Finally, Section~\ref{sect:discussion} covers the related work, the future work, and the conclusion.

>>>>

An important goal in Requirements Engineering (RE) is to produce a consistent and comprehensive set of requirements covering different aspects of the system, such as general functional requirements, operational environment constraints, and so-called non-functional requirements such as security and performance. Among the ``early-phase'' requirements engineering activities are those that consider how the intended system should meet organizational goals, why it is needed, what alternatives may exist, what the implications of the alternatives are for different stakeholders, and how the interests and concerns of stakeholders might be addressed~\cite{yu1997towards}. These activities fall under the umbrella of goal modeling. There are a large number of established RE methods using goal models in the early stage of requirements analysis (overviews can be found in~\cite{kavakliL05,van2001goal}). Several goal modeling languages have been developed in the last two decades as well. The most popular ones include $i*$~\cite{yu1997towards}, Keep All Objects Satisfied (KAOS)~\cite{van2008requirements}, the NFR framework~\cite{chung2012non}, \textsc{Tropos}~\cite{giorgini2005goal}, the Business Intelligence Model (BIM)~\cite{horkoff2014strategic}, and the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}.

@@

\begin{enumerate}
\item 
The argumentation techniques must capture the actual discussions of the stakeholders or designers in the early requirements engineering phase.
\item 
The framework must have formal traceability links between elements of the goal model and their underlying arguments.
\item 
Using these traceability links, it must be possible to compute the effect of changes in the underlying arguments about the goal model, and vice versa.
\item 
A process explaining how practitioners can develop models in the framework must be identified.
\item 
The framework must have tool support.
\end{enumerate}

@@

In order to support practitioners in using the RationalGRL framework, we propose the RationalGRL development process, which explains steps for  developing goal models and posing arguments based on schemes in an integrated way. We implement a web-based tool for the RationalGRL framework as well, which works on any modern browser and is developed in Javascript.\footnote{The tool and source code can be found at \rationalgrlurl.} The tool is open-source, and designed such that new argument schemes and critical questions can be added easily.

The rest of this article is organized as follows. Section~\ref{sect:background} introduces our running example and briefly discusses the basics of GRL and argumentation about goals using argument schemes. Section~\ref{sect:gmas} contains the case study and an explanation of how we obtained an initial set of argument schemes and critical questions by coding transcripts from discussions about an information system. Section~\ref{sect:overview} provides an overview of the RationalGRL framework and the RationalGRL metamodel, and examples from the case study that illustrate the framework and its language. In Section~\ref{sect:formalframework}, we provide formal semantics for GRL and RationalGRL, show how RationalGRL models can be translated to GRL models and vice versa, and we develop various algorithms that change a RationalGRL model according to an argument scheme or a critical question. Section~\ref{sect:methodology+tool} discusses the RationalGRL development process and explains various features of the tool we developed. Finally, Section~\ref{sect:discussion} covers the related work, the future work, and the conclusion.
